---
title: "HRRR smoke for 1000 and 6000 ft"
output: html_document
---

Extract **1000 ft** and **6000 ft AGL smoke mass density** (MASSDEN) from HRRR
at weather station locations.

Output CSV columns: `datetime`, `siteID`, `smoke1000`, `smoke6000` (units: μg/m³)

```{r setup, message=FALSE, warning=FALSE}

library(httr)
library(terra)
library(tidyverse)
library(tictoc)

```

# Setup

```{r config}

stations <- "/Users/megansears/Documents/ARS/HRRR/AmeriFlux_Sites_SKP.csv" # needs lat, lon, siteID
start_date <- ymd_h("2021-05-31 00") # UTC
end_date <- ymd_h("2021-10-01 12") # UTC,
fxx<- c(0) # forecast hour 0 means model's estimate, not a forecast
interval_hr <- 1 # timestep
output <- "smoke_test.csv"

# HRRR hybrid sigma level numbers (1=top, 50=near surface)
# From HRRR published eta level table:
#   Level 42 ≈ 306 m AGL (~1000 ft)
#   Level 29 ≈ 1863 m AGL (~6000 ft)
level1000 <- 5
level6000 <- 11

# Units are in kg/m3

```

# Load Stations

```{r load-stations}

stations_df <- read_csv(stations)

```

# Helper Functions

```{r helpers}

# Build the HRRR AWS S3 URL for a given date, forecast hour, and product
hrrr_url <- function(date, fxx, product = "nat") {
  date_str <- format(date, "%Y%m%d")
  hour_str <- format(date, "%H")
  fxx_str  <- sprintf("%02d", fxx)
  paste0(
    "https://noaa-hrrr-bdp-pds.s3.amazonaws.com/hrrr.", date_str,
    "/conus/hrrr.t", hour_str, "z.wrf", product, "f", fxx_str, ".grib2"
  )
}

# Download and parse the HRRR .idx file into a data frame
parse_idx <- function(url) {
  idx_url <- paste0(url, ".idx")
  lines <- tryCatch(readLines(idx_url, warn = FALSE), error = function(e) NULL)
  if (is.null(lines)) return(NULL)

  # idx format: msg_num:byte_start:d=YYYYMMDDHH:variable:level:fcst_type:
  parts <- strsplit(lines, ":")
  df <- data.frame(
    byte_start = as.numeric(sapply(parts, `[`, 2)),
    variable   = trimws(sapply(parts, `[`, 4)),
    level      = trimws(sapply(parts, `[`, 5)),
    stringsAsFactors = FALSE
  )
  # End byte = start of next message minus 1 (last message reads to EOF)
  df$byte_end <- c(df$byte_start[-1] - 1, NA)
  df
}

# Download a single GRIB2 message via HTTP byte-range request
download_grib_message <- function(url, byte_start, byte_end) {
  range_str <- if (!is.na(byte_end)) {
    paste0("bytes=", byte_start, "-", byte_end)
  } else {
    paste0("bytes=", byte_start, "-")
  }
  tmpfile <- tempfile(fileext = ".grib2")
  resp <- GET(url, add_headers(Range = range_str), write_disk(tmpfile, overwrite = TRUE))
  stop_for_status(resp)
  tmpfile
}

# Fetch a variable at one hybrid level; returns a terra SpatRaster
fetch_level <- function(url, idx, variable, hybrid_level) {
  target <- paste0(hybrid_level, " hybrid level")
  row <- idx[grepl(variable, idx$variable, ignore.case = TRUE) &
               idx$level == target, ]

  if (nrow(row) == 0) {
    warning(variable, " at '", target, "' not found in index")
    return(NULL)
  }

  tmpfile <- download_grib_message(url, row$byte_start[1], row$byte_end[1])
  terra::rast(tmpfile)
}

# Extract raster values at station locations (nearest neighbor)
extract_at_stations <- function(r, sv) {
  vals <- terra::extract(r, sv)
  vals[[2]]  # first col is ID, second is the value
}

```

# Verify Hybrid Level Heights

Downloads geopotential height (HGT) at all 50 hybrid levels and terrain height
from the surface product, then calculates AGL height at the center of the HRRR
grid. Use this to confirm `level1000` and `level6000` are set correctly before
running the full extraction.

```{r verify, message=FALSE}

url_nat <- hrrr_url(start_date, fxx[1], product = "nat")
url_sfc <- hrrr_url(start_date, fxx[1], product = "sfc")
idx_nat  <- parse_idx(url_nat)
idx_sfc  <- parse_idx(url_sfc)

# Check what height variable is available in the nat file
hgt_var <- unique(idx_nat$variable[grepl("^HGT$", idx_nat$variable, ignore.case = TRUE)])
if (length(hgt_var) == 0) {
  hgt_var <- unique(idx_nat$variable[grepl("HGT|gh|HGHT", idx_nat$variable)])
}
cat("Height variable found in nat idx:", paste(hgt_var, collapse = ", "), "\n\n")

# Download terrain height from sfc product
terrain_var <- unique(idx_sfc$variable[grepl("^HGT$", idx_sfc$variable, ignore.case = TRUE)])[1]
terrain_row  <- idx_sfc[idx_sfc$variable == terrain_var &
                          grepl("surface", idx_sfc$level, ignore.case = TRUE), ]

if (nrow(terrain_row) > 0) {
  r_terrain  <- terra::rast(download_grib_message(url_sfc, terrain_row$byte_start[1],
                                                   terrain_row$byte_end[1]))
  # Sample at grid center
  cx <- ncol(r_terrain) %/% 2
  cy <- nrow(r_terrain) %/% 2
  terrain_m <- as.numeric(r_terrain[cy, cx])
  cat("Terrain height at grid center:", round(terrain_m, 1), "m\n\n")
} else {
  terrain_m <- 0
  cat("Could not find terrain height — using 0 m (sea level assumption)\n\n")
}

# Download HGT at each hybrid level and compute AGL height at grid center
if (length(hgt_var) > 0) {
  agl_results <- data.frame(hybrid_level = integer(), height_agl_m = numeric(),
                             height_agl_ft = numeric())

  for (lvl in 1:50) {
    r_hgt <- fetch_level(url_nat, idx_nat, hgt_var[1], lvl)
    if (is.null(r_hgt)) next

    cx <- ncol(r_hgt) %/% 2
    cy <- nrow(r_hgt) %/% 2
    hgt_m   <- as.numeric(r_hgt[cy, cx])
    agl_m   <- hgt_m - terrain_m
    agl_ft  <- agl_m * 3.28084

    agl_results <- rbind(agl_results,
                         data.frame(hybrid_level = lvl,
                                    height_agl_m  = round(agl_m, 1),
                                    height_agl_ft = round(agl_ft, 0)))
  }

  cat("Hybrid level AGL heights at grid center:\n")
  print(agl_results)

  # Find levels closest to targets
  closest_1000 <- agl_results[which.min(abs(agl_results$height_agl_m - 305)), ]
  closest_6000 <- agl_results[which.min(abs(agl_results$height_agl_m - 1829)), ]

  cat("\nClosest level to 1000 ft (305 m):\n"); print(closest_1000)
  cat("\nClosest level to 6000 ft (1829 m):\n"); print(closest_6000)
  cat("\nUpdate level1000 and level6000 in the config chunk if needed.\n")

} else {
  cat("HGT not found in nat idx — cannot verify level heights.\n")
  cat("Proceeding with level1000 =", level1000, "and level6000 =", level6000, "\n")
}

```

# (Optional) Inspect Target Levels

Quick check that MASSDEN downloads correctly for the two target levels.

```{r inspect, message=FALSE}

url_test <- hrrr_url(start_date, fxx[1])
idx_test  <- parse_idx(url_test)

for (lvl_info in list(c("1000 ft", level1000), c("6000 ft", level6000))) {
  label <- lvl_info[1]
  level <- as.integer(lvl_info[2])
  r <- fetch_level(url_test, idx_test, "MASSDEN", level)
  if (!is.null(r)) {
    cat(label, "-> hybrid level", level, " | dims:", nrow(r), "x", ncol(r),
        " | layer name:", names(r), "\n")
  }
}

```

# Extract Time Series

```{r extract, message=FALSE}

tic()
sv <- vect(stations_df, geom = c("lon", "lat"), crs = "EPSG:4326")

all_records <- list()

cycle_times <- seq(start_date, end_date, by = paste(interval_hr, "hours"))

for (dt in cycle_times) {
  dt <- as_datetime(dt)

  for (f in fxx) {
    cat("Fetching", format(dt, "%Y-%m-%d %H:%M"), "UTC  fxx=", sprintf("%02d", f), "h ...")

    url <- hrrr_url(dt, f)
    idx <- parse_idx(url)

    if (is.null(idx)) {
      cat(" [WARN] could not read index\n")
      next
    }

    r_1000 <- fetch_level(url, idx, "MASSDEN", level1000)
    r_6000 <- fetch_level(url, idx, "MASSDEN", level6000)

    if (is.null(r_1000) || is.null(r_6000)) {
      cat(" [WARN] one or both levels missing\n")
      next
    }

    sv_proj    <- project(sv, crs(r_1000))
    smoke_1000 <- extract_at_stations(r_1000, sv_proj)
    smoke_6000 <- extract_at_stations(r_6000, sv_proj)

    valid_time <- dt + hours(f)

    records <- data.frame(
      datetime  = format(valid_time, "%Y-%m-%d %H:%M:%S"),
      siteID    = stations_df$siteID,
      smoke1000 = round(smoke_1000, 10),
      smoke6000 = round(smoke_6000, 10),
      stringsAsFactors = FALSE
    )
    all_records <- append(all_records, list(records))
    cat("", nrow(records), "rows\n")
  }
}

df_out <- bind_rows(all_records) |>
  arrange(datetime, siteID)

cat("\nTotal rows:", nrow(df_out), "\n")

toc()

```

# Save & Preview

```{r save}

write_csv(df_out, output)

head(df_out, 20)

```
