---
title: "HRRR smoke for 1000 and 6000 ft"
output: html_document
---

Extract **1000 ft** and **6000 ft AGL smoke mass density** (MASSDEN) from HRRR
at weather station locations.

Output CSV columns: `datetime_utc`, `siteID`, `smoke1000`, `smoke6000` (units: μg/m³)

```{r setup, message=FALSE, warning=FALSE}

library(httr)
library(terra)
library(tidyverse)
library(tictoc)

```

# Setup

```{r config}

stations <- "/Users/megansears/Documents/ARS/HRRR/AmeriFlux_Sites_SKP.csv" # needs lat, lon, siteID

start_date <- ymd_h("2025-05-31 00") # UTC
end_date <- ymd_h("2025-10-01 23") # UTC,

fxx<- c(0) # forecast hour 0 means model's estimate, not a forecast

interval_hr <- 1 # timestep

output <- "smoke_2025_test.csv"

```

From brian.d.jamison@noaa.gov: 

git clone https://github.com/NOAA-GSL/pygraf
This package plots many variables, but 1000ft and 6000ft smoke are listed in the file:
adb_graphics/default_specs.yml

They are under the heading "trc1" and sub-headings "1000ft" and "6000ft".  
The "vertical_index" gives the index used for the level, 
so index 4 is used for 1000ft and index 11 is used for 6000ft.


```{r config}

# Brian calls out index 4 and 11 as 1000 and 6000 ft smoke variables
# but in that code there the values start at 0
# in the index file grib2 we are using here, the values start at 1
# so using index 5 and 12 to get 1000 and 6000 ft smoke vars
level1000 <- 5
level6000 <- 12

```

# Load Stations

```{r load-stations}

stations_df <- read_csv(stations)

```

# Helper Functions

```{r helpers}

# Build the HRRR AWS S3 URL for a given date, forecast hour, and product
hrrr_url <- function(date, fxx, product = "nat") {
  date_str <- format(date, "%Y%m%d")
  hour_str <- format(date, "%H")
  fxx_str  <- sprintf("%02d", fxx)
  paste0(
    "https://noaa-hrrr-bdp-pds.s3.amazonaws.com/hrrr.", date_str,
    "/conus/hrrr.t", hour_str, "z.wrf", product, "f", fxx_str, ".grib2"
  )
}

# Download and parse the HRRR .idx file into a data frame
# The idx is the parsed HRRR index that tells the code where in the GRIB2
# each variable is, which enables a targeted byte-range HTTP
# instead of pulling the whole grib2
parse_idx <- function(url) {
  idx_url <- paste0(url, ".idx")
  lines <- tryCatch(readLines(idx_url, warn = FALSE), error = function(e) NULL)
  if (is.null(lines)) return(NULL)

  # idx format: msg_num:byte_start:d=YYYYMMDDHH:variable:level:fcst_type:
  parts <- str_split(lines, ":")
  tibble(
    byte_start = map_dbl(parts, ~ as.numeric(.x[2])),
    variable   = map_chr(parts, ~ str_trim(.x[4])),
    level      = map_chr(parts, ~ str_trim(.x[5]))
  ) %>%
    mutate(byte_end = c(byte_start[-1] - 1, NA))
}

################################################################################

# Download a single GRIB2 via HTTP byte-range request
download_grib_message <- function(url, byte_start, byte_end) {
  range_str <- if (!is.na(byte_end)) {
    paste0("bytes=", byte_start, "-", byte_end)
  } else {
    paste0("bytes=", byte_start, "-")
  }
  tmpfile <- tempfile(fileext = ".grib2")
  resp <- GET(url, add_headers(Range = range_str), 
              write_disk(tmpfile, overwrite = TRUE))
  stop_for_status(resp)
  tmpfile
}

################################################################################

# Fetch a variable at one hybrid level; returns a terra SpatRaster
fetch_level <- function(url, idx, variable, hybrid_level) {
  target <- paste0(hybrid_level, " hybrid level")
  row <- idx %>%
    filter(str_detect(variable, regex(.env$variable, ignore_case = TRUE)),
           level == target)

  if (nrow(row) == 0) {
    warning(variable, " at '", target, "' not found in index")
    return(NULL)
  }

  tmpfile <- download_grib_message(url, row$byte_start[1], row$byte_end[1])
  r <- terra::rast(tmpfile)

  # GRIB2 packing varies by HRRR vintage — GDAL auto-scales some files to ug/m3,
  # others return raw kg/m3. Metadata is not exposed, so detect by value range:
  # kg/m3 values are always < 1e-5; ug/m3 values are >> 1e-5.
  r_max <- global(r, "max", na.rm = TRUE)[[1]]
  if (!is.na(r_max) && r_max < 1e-5) r <- r * 1e9

  r
}

################################################################################

# Extract raster values at station locations
extract_at_stations <- function(r, sv) {
  terra::extract(r, sv) %>% 
    pull(2)  # col 1 is ID, col 2 is the value
}

```

# Inspect Target Levels

Quick check that MASSDEN downloads correctly for the two target levels.
Also used this to verify units.

```{r inspect, message=FALSE}

url_test <- hrrr_url(start_date, fxx[1])
idx_test  <- parse_idx(url_test)

tibble(
  label = c("1000 ft", "6000 ft"),
  level = c(level1000, level6000)
) %>%
  pwalk(\(label, level) {
    # Download the GRIB2 message directly so we can inspect via describe()
    target <- paste0(level, " hybrid level")
    row <- idx_test %>%
      filter(str_detect(variable, "MASSDEN"), level == target)

    if (nrow(row) == 0) { cat(label, ": not found\n"); return() }

    tmpfile <- download_grib_message(url_test, row$byte_start[1], row$byte_end[1])
    r <- terra::rast(tmpfile)

    cat(label, "-> hybrid level", level, "\n")
    cat("  Layer name:", names(r), "\n")

    # GDAL full metadata — look for GRIB_DECIMAL_SCALE_FACTOR or similar
    cat("  describe() metadata:\n")
    print(terra::describe(tmpfile, meta = TRUE))

    cx <- ncol(r) %/% 2; cy <- nrow(r) %/% 2
    cat("  Center pixel:", round(as.numeric(r[cy, cx]), 12), "\n")
    cat("  Raster min/max:", round(minmax(r), 12), "\n\n")
  })

```

# Extract Time Series

```{r extract, message=FALSE}

tic()

# Convert station lat/lon to spatial vector in WGS84; reprojected to HRRR CRS each timestep
sv <- vect(stations_df, 
           geom = c("lon", "lat"), 
           crs = "EPSG:4326")

# One model run (dt x fxx combination) -> tibble of station smoke values, or NULL on failure
fetch_smoke <- function(dt, f) {
  cat("Fetching", format(dt, "%Y-%m-%d %H:%M"), "UTC  fxx=", sprintf("%02d", f), "h ...")

  # Build S3 URL and parse .idx to get byte offsets for each variable/level
  url <- hrrr_url(dt, f)
  idx <- parse_idx(url)

  if (is.null(idx)) { cat(" [WARN] could not read index\n");        return(NULL) }

  # Byte-range download — only fetches the two target MASSDEN levels, not the full GRIB2
  r_1000 <- fetch_level(url, idx, "MASSDEN", level1000)
  r_6000 <- fetch_level(url, idx, "MASSDEN", level6000)

  if (is.null(r_1000) || is.null(r_6000)) { cat(" [WARN] one or both levels missing\n"); return(NULL) }

  # Reproject stations into HRRR's Lambert Conformal Conic CRS before extracting
  sv_proj <- project(sv, crs(r_1000))
  valid_time <- dt + hours(f)  # fxx=0: valid time == model run time (analysis, not forecast)

  out <- tibble(
    datetime = format(valid_time, "%Y-%m-%d %H:%M:%S"),
    siteID = stations_df$siteID,
    smoke1000 = round(extract_at_stations(r_1000, sv_proj), 4),  # ug/m3 (normalized in fetch_level)
    smoke6000 = round(extract_at_stations(r_6000, sv_proj), 4),  # ug/m3
  )
  cat("", nrow(out), "rows\n")
  out
}

# All dt x fxx combinations -> fetch, drop failures, combine
df_out <- expand_grid(
    dt = as_datetime(seq(start_date, 
                         end_date, 
                         by = paste(interval_hr, "hours"))),
    f  = fxx
  ) %>%
  pmap(fetch_smoke) %>%
  compact() %>%      # drop NULLs from failed/missing timesteps
  list_rbind() %>%
  arrange(datetime, siteID)

cat("\nTotal rows:", nrow(df_out), "\n")

toc()

```

# Save & Preview

```{r save}

# Values are normalized to ug/m3 in fetch_level; rename columns accordingly
df_out <- df_out %>%
  rename(smoke1000_ugm3 = smoke1000,
         smoke6000_ugm3 = smoke6000)

summary(df_out)

write_csv(df_out, output)

```

# Testing

```{r}

test <- df_out %>%
  filter(datetime == ymd_hms("2025-06-03 12:00:00",
                             tz = 'UTC'))

locs_test <- stations_df %>%
  left_join(test, by = "siteID") %>%
  vect(geom = c("lon", "lat"), crs = "EPSG:4326")

mapview(locs_test, zcol = "smoke1000_ugm3")

# try to match scale and colors to line up with HRRR smoke graphics
clevs <- c(1, 2, 4, 6, 8, 12, 16, 20, 25, 30, 40, 60, 100, 200)

smoke_cols <- c(
  "#f7fbff", "#deebf7", "#9ecae1", "#4292c6", "#2171b5", "#08519c",
  "#41ab5d", "#addd8e", "#ffffb2", "#fecc5c", "#fd8d3c", "#f03b20",
  "#bd0026", "#dd3497"
)

stations_df %>%
  left_join(test, by = "siteID") %>%
  ggplot(aes(x = lon, y = lat, fill = smoke1000_ugm3)) +
  borders("state", colour = "gray70", fill = "white") +
  geom_point(size = 3, shape = 21, color = "gray30", stroke = 0.4) +
  scale_fill_stepsn(
    colors  = smoke_cols,
    breaks  = clevs,
    limits  = c(0, 250),
    name    = "Smoke 1000ft\n(μg/m³)"
  ) +
  guides(fill = guide_colorsteps(even.steps = TRUE, 
                                 barheight = unit(8, "cm"),
                                 show.limits = TRUE)) +
  coord_fixed(1.3) +
  theme_minimal() +
  labs(title = "HRRR Smoke at 1000ft AGL",
       subtitle = "2025-06-04 06:00 UTC",
       x = NULL, y = NULL)

```

